---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file, then run rmarkdown::render("README.Rmd"). -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.path = "README-",
  dpi = 150,
  fig.width = 8,
  fig.height = 3.6,
  out.width = "100%"
)
```

<!-- badges: start -->
[![R-CMD-check](https://github.com/Ddey07/SGCTools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Ddey07/SGCTools/actions/workflows/R-CMD-check.yaml)
[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
<!-- badges: end -->

# SGCTools

> Latent correlation, covariance estimation, and functional PCA for **mixed-type data** — continuous, truncated, ordinal, and binary — via the Semiparametric Gaussian Copula (SGC).

Most multivariate and functional methods assume Gaussian (or at least
continuous) measurements. Real data are rarely so tidy: a single study may mix
continuous biomarkers, zero-inflated (truncated) activity counts, ordinal
questionnaire items, and binary indicators. `SGCTools` models all of these
through a shared **latent Gaussian process**, so you can estimate correlations,
run PCA, fit regressions, and do functional PCA on mixed-type data with one
consistent framework.

The methods are described in:

- Dey, D., Ghosal, R., Merikangas, K., & Zipunnikov, V. (2024). *Functional
  Principal Component Analysis for Continuous Non-Gaussian, Truncated, and
  Discrete Functional Data.* **Statistics in Medicine**, 43, 5431–5445.
  <https://doi.org/10.1002/sim.10240>
- Dey, D., & Zipunnikov, V. (2022). *Semiparametric Gaussian Copula Regression
  modeling for Mixed Data Types (SGCRM).* <https://arxiv.org/abs/2205.06868>

## What you can do

- **Latent correlation** for any mix of continuous / truncated / ordinal /
  binary variables (`fromXtoRMixed`).
- **Fast Kendall's τ** with missing-data support, in `O(n log n)`
  (`Kendall_mixed`, backed by C++).
- **Latent-space regression** with asymptotic standard errors and p-values
  (`sgclm`).
- **Predictions (BLUPs) of the latent variables** from observed mixed data
  (`getLatentPreds`).
- **Functional PCA** of a continuous / truncated / ordinal / binary functional
  process (`fpca.sgc.lat`).

| Function          | Purpose                                                        |
|-------------------|----------------------------------------------------------------|
| `fromXtoRMixed`   | Estimate the latent correlation matrix from mixed-type data    |
| `Kendall_mixed`   | Fast Kendall's τ (handles missing values)                      |
| `sgclm`           | Latent semiparametric Gaussian copula regression               |
| `getLatentPreds`  | BLUPs of the latent Gaussian variables                         |
| `fpca.sgc.lat`    | Covariance estimation + FPCA for mixed-type functional data    |

## Installation

```r
# install.packages("devtools")
devtools::install_github("Ddey07/SGCTools")
```

## Example 1 — Functional PCA of binary data

We simulate a latent Gaussian process with a Matérn covariance, threshold it to
get a **binary** functional process, and recover the latent covariance surface
and its eigenstructure with `fpca.sgc.lat`.

```{r fpca-example}
library(SGCTools)

matern <- function(u, phi, kappa) {
  uphi <- u / phi
  uphi <- ifelse(u > 0,
                 (2^(-(kappa - 1)) / gamma(kappa)) * (uphi^kappa) *
                   besselK(x = uphi, nu = kappa),
                 1)
  uphi[u > 600 * phi] <- 0
  uphi
}

set.seed(1)
n <- 1000; m <- 15
tp <- seq(0, 1, length = m)
d  <- abs(outer(tp, tp, "-"))            # distance matrix |s - t|
C_true <- matern(d, phi = 1/2, kappa = 3.5)

y <- mvtnorm::rmvnorm(n, sigma = C_true) # latent Gaussian process
z <- (y > 0.5) * 1                       # observed *binary* process

ff <- fpca.sgc.lat(X = z, type = "bin", argvals = tp, df = 4)
```

The SGC estimate recovers the latent correlation surface from binary
observations alone:

```{r fpca-surfaces, fig.height = 3.6}
pal <- hcl.colors(64, "YlOrRd", rev = TRUE)
zl  <- range(c(C_true, as.matrix(ff$cov)))

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
image(tp, tp, C_true, col = pal, zlim = zl,
      xlab = "s", ylab = "t", main = "True correlation  C(s, t)")
image(tp, tp, as.matrix(ff$cov), col = pal, zlim = zl,
      xlab = "s", ylab = "t", main = "SGC estimate (from binary data)")
```

...along with interpretable eigenfunctions and a fast-decaying eigenvalue
spectrum:

```{r fpca-eigen, fig.height = 3.4}
cols <- hcl.colors(3, "Dark 3")
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

matplot(tp, ff$efunctions[, 1:3], type = "l", lwd = 2, lty = 1, col = cols,
        xlab = "t", ylab = "eigenfunction", main = "Leading eigenfunctions")
abline(h = 0, col = "grey80")
legend("topright", paste0("PC", 1:3), col = cols, lwd = 2, bty = "n")

barplot(ff$evalues, names.arg = paste0("PC", seq_along(ff$evalues)),
        col = "steelblue", border = NA, ylab = "eigenvalue",
        main = "Eigenvalue spectrum")
```

## Example 2 — Latent correlation for mixed-type data

The real strength of SGC is recovering a single latent correlation structure
across variables of *different* types. Here six variables share a latent
correlation `S`, but we only observe them as continuous, log-transformed
continuous, binary, ordinal, and truncated measurements.

```{r mixed-example}
set.seed(2)
p <- 6
S <- clusterGeneration::rcorrmatrix(p)        # true latent correlation
L <- mvtnorm::rmvnorm(1000, sigma = S)        # latent Gaussian

X <- data.frame(
  continuous = L[, 1],
  lognormal  = exp(L[, 2]),                          # monotone transform
  binary     = as.numeric(L[, 3] > 0),
  ordinal    = as.numeric(cut(L[, 4], c(-Inf, -0.5, 0.5, Inf))) - 1,
  truncated  = ifelse(L[, 5] > 0, L[, 5], 0),        # point mass at 0
  binary2    = as.numeric(L[, 6] > 0.3)
)

Rhat <- fromXtoRMixed(X)$hatR                  # estimated latent correlation
```

Despite the mixed types and nonlinear marginals, the estimated latent
correlation closely matches the truth:

```{r mixed-heatmap, fig.height = 4}
corr_heat <- function(M, title) {
  pal <- hcl.colors(64, "Blue-Red 3")
  image(1:p, 1:p, t(M[p:1, ]), col = pal, zlim = c(-1, 1),
        axes = FALSE, xlab = "", ylab = "", main = title)
  axis(1, 1:p, colnames(X), las = 2, cex.axis = 0.7, tick = FALSE)
  axis(2, 1:p, rev(colnames(X)), las = 2, cex.axis = 0.7, tick = FALSE)
}
par(mfrow = c(1, 2), mar = c(6, 6, 3, 1))
corr_heat(S,    "True latent correlation")
corr_heat(Rhat, "SGC estimate")
```

You can then use the estimated structure directly — e.g. PCA on the latent
correlation, or a latent regression with `sgclm`:

```{r sgclm}
fit <- sgclm(binary ~ continuous + ordinal + truncated, data = X)
fit$coef
```

## Citation

If you use `SGCTools`, please cite:

> Dey, D., Ghosal, R., Merikangas, K., & Zipunnikov, V. (2024). Functional
> Principal Component Analysis for Continuous Non-Gaussian, Truncated, and
> Discrete Functional Data. *Statistics in Medicine*, 43, 5431–5445.
> doi:10.1002/sim.10240

## License

GPL (>= 3).
